---
title: Homework Assignment 1 EDS 223
format: 
  pdf:
    theme: Flatly   
editor: visual
---

---
title: Environmental Injustice in the Kansas City Metropolitan Area
author: William Mullins
date: 10-06-2025
---

### Reading in Packages and Dataset

------------------------------------------------------------------------

```{r, message=FALSE}
# Importing Required Packages
library(tidyverse)
library(sf)
library(tmap)
library(here)
```

```{r, echo=TRUE, results='hide'}
# Read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) 
```

### Data Preparation

------------------------------------------------------------------------

```{r}
# Read in Kansas and Missouri data sets
# Due to overlapping county names this must be done seperatly
kansas <- ejscreen %>%
  dplyr::filter(ST_ABBREV == c('KS')) 
missouri <- ejscreen %>%
  dplyr::filter(ST_ABBREV == c('MO')) 

# Read in counties located in the Kansas City Metropolitan Area
county_map_kansas <- kansas %>%
      filter(CNTY_NAME %in% c("Johnson County", 
                              "Wyandotte County", 
                              "Leavenworth County"))
county_map_missouri <- missouri %>%
      filter(CNTY_NAME %in% c("Jackson County", 
                              "Platte County", 
                              "Clay County", 
                              "Cass County"))

# Merge the two data sets together by binding rows
kcmo_counties <- bind_rows(county_map_kansas, county_map_missouri)

# Specify the area the map will display 
custom_bbox <- c(xmin = -94.83, ymin = 38.875, xmax = -94.35, ymax = 39.25)
```

### Figure 1. The Prevalence of Low Income Households and the Release of Airborne Toxins

------------------------------------------------------------------------

```{r, message=FALSE, warning=FALSE}
# Bivariate plot of airborne toxin release and percent of households in low income
tm_shape(kcmo_counties, bbox = custom_bbox) +
   tm_graticules(x = c(-94.80, -94.70, -94.60, -94.50, -94.40, -94.30), 
                 crs = 4326,
                 labels.show = TRUE) + # Show coordinates on x and y of map
   tm_polygons(fill = tm_vars(c("RSEI_AIR", "LOWINCPCT"), multivariate = TRUE),
    fill.scale = tm_scale_bivariate(
      values = "brewer.seqseq2",
      scale1 = tm_scale_intervals(
               breaks = c(5, 25, 100, 500, 2000, 10000, 50000, 250000, 1250000),
                labels = c("(5", "25", "100", "500", "2,000", 
                           "10,000", "50,000", "250,000", "1,250,000")),
              scale2 = tm_scale_intervals(
                breaks = c(0, 0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.7, 0.8, 0.9, 1), 
                labels = c("0", "10", "20", "30", "40", "50", "60", "70", "80", "90", "100"))),
            # Format the legend and give titles
            fill.legend = tm_legend_bivariate(
              orientation = "landscape",
              xlab = "Percent Low Income (10%)", 
              ylab = "Toxic Releases to Air µg/m (m\u00B3)",
              xlab.size=2.3,
              ylab.size=2.3,
              item.height = 1.5,
              item.width = 1.5,
              reverse = FALSE
            )
    ) +
   tm_layout(main.title = "Airborn Toxin Release and % Low Income \n in the Kansas City Metropolitan Area",
            main.title.fontface = "bold", 
            main.title.position = "center",
            legend.title.fontfamily = "monospace",
            legend.text.size = 2)
```

### Figure 2. The Prevalence of People of Color and the Release of Airborne Toxins

------------------------------------------------------------------------

```{r, message=FALSE, warning=FALSE}
# Bivariate plot of the percent of people of color and the release of airborn3 toxins
tm_shape(kcmo_counties, bbox = custom_bbox) +
   tm_graticules(x = c(-94.80, -94.70, -94.60, -94.50, -94.40, -94.30),
                 crs = 4326,
                 labels.show = TRUE) + 
   tm_polygons(fill = tm_vars(c("RSEI_AIR", "PEOPCOLORPCT"), multivariate = TRUE),
    fill.scale = tm_scale_bivariate(
      values = "brewer.seqseq2",
      scale1 = tm_scale_intervals(
                breaks = c(5, 25, 100, 500, 2000, 10000, 50000, 250000, 1250000),
                labels = c("(5", "25", "100", "500", "2,000", 
                           "10,000", "50,000", "250,000", "1,250,000")),
              scale2 = tm_scale_intervals(
                breaks = c(0.10, 0.20, 0.30, 0.40, 0.50, 0.60, 0.7, 0.8, 0.9, 1), 
                labels = c("10", "20", "30", "40", "50", "60", "70", "80", "90", "100")
              )
            ),
            # Format the legend and give titles
            fill.legend = tm_legend_bivariate(
              orientation = "landscape",
              xlab = "Percent People of Color (10%)", 
              ylab = "Toxic Releases to Air", 
              xlab.size=2.3,
              ylab.size=2.3,
              item.height = 1.5,
              item.width = 1.5,
              reverse = FALSE)) +
   tm_layout(main.title = "Release of Airborn Toxins and % People of Color \n in the Kansas City Metropolitan Area",
            main.title.fontface = "bold", 
            main.title.position = "center",
            legend.title.fontfamily = "monospace",
            legend.text.size = 2)
```

## Discussion

------------------------------------------------------------------------

These maps reveal that communities in the Kansas City area with higher percentages of people of color and/or low-income residents are constantly located in the same areas with the highest levels of airborne toxic releases. The overlap is prominent, with nearly every block with a high proportion of either group being the darkest-shaded zones on the map, with a notable lack of blue colored areas. Interestingly, both maps also show the state line between Kansas and Missouri despite not being explicitly drawn in. This indicates that communities on the Missouri side may be more exposed to harmful airborne chemicals than those in Kansas.
